### 2023-02-08 paper-thesis/chapters/10-coc.tex
More writing on coc

An example program is the following, which defines a polymorphic identity function and applies it to a function:

@@

@@

In Statix, all this information can be stored in a \emph{scope graph}~\cite{scope_graphs}, as explained in chapter \ref{chap:bg-spoofax}. We only use a single type of edge, called \verb|P| (parent) edges. We also only have a single relation, called \verb|name|. This name stores a \verb|NameEntry|, which can be either a \verb|NType|, which stores the type of a name, or a \verb|NSubst|, which stores a name that has been substituted with a value. The Statix definition of these concepts is given below:
\begin{lstlisting}
constructors
    NameType : Expr -> NameEntry
    NameSubst : scope * Expr -> NameEntry
relations
    name : ID -> NameEntry
\end{lstlisting}

@@

We implemented beta reduction using a Krivine abstract machine~\cite{krivine}. The machine can head evaluate lambda expressions with a call-by-name semantics. It works by keeping a stack of all arguments that have not been applied yet. This turned out to be the more natural way of expressing this compared to substitution-based evaluation relation, which is an alternative we will discuss in section \ref{sec:coc-subst}.

In conventional dependently typed languages, evaluation is often done using De Bruijn indices. However, we chose to use names rather than De Bruijn indices, because scope graphs work based on names. Using De Bruijn indices would also prevent us from using editor services that rely on \verb|.ref| annotations (which are Spoofax annotations that declare one name as being a use of another name that is a definition), such as renaming.

We need to define multiple predicates that will be used later for type checking. First, the primary predicate is \verb|betaReduceHead|, that takes a scoped expression and a stack of applications, and returns a head-normal expression. The scope acts as the environment from~\cite{krivine}, using \verb|NSubst| to store substitutions. All rules for \verb|betaReduceHead| are given in figure \ref{fig:beta-head-reduce-rules}. We use the syntax $\bhr{\scope{s_1}{e_1}}{\overline{p}}{\scope{s_2}{e_2}}$ to express \verb|betaReduceHead((s1, e1), ps) == (s2, e2)|. The \verb|p| in this definition is the argument stack of the Krivine machine. Figure \ref{fig:beta-head-reduce-rules} contains the rules necessary for beta head reduction of the language. One predicate that is used for this is the \verb|rebuild| predicate, which takes a scoped expression and the stack of arguments (of the Krivine machine state) and converts it to an expression by adding \verb|FnDestruct|s. Its signature is:
\begin{lstlisting}
rebuild : (scope * Expr) * list((scope * Expr)) -> (scope * Expr)
\end{lstlisting}

@@

The inference rules in figure \ref{fig:type-check-rules} can be directly translated to Statix rules. For example, the rule for \verb|Let| bindings is expressed like this in Statix:
\begin{lstlisting}
typeOfExpr(s, Let(n, v, b)) = typeOfExpr(s', b) :-
    typeOfExpr(s, v) == vt, sPutSubst(s, n, (s, v)) == s'.
\end{lstlisting}

@@

Although these rules look cleaner, they are more complicated to implement, requiring a separate relation to check if \verb|f| is a \verb|FnConstruct| or something else. For the pure calculus of constructions, this still works quite well but when adding inductive datatypes (chapter \ref{ch:datatypes}) this quickly becomes a mess.

>>>>

An example program is the following, which defines a polymorphic identity function (an identity function that is generic over its type) and applies it to a function:

@@

The type that the function is generic over needs to be explicitly specified. In most languages, generics are inferred, this inference will also be possible in this language after implementing inference in chapter~\ref{ch:inference}.

@@

In Statix, all this information can be stored in a \emph{scope graph}~\cite{scope_graphs}, as explained in chapter \ref{chap:bg-spoofax}. We only use a single type of edge, called \verb|P| (parent) edges. We also only have a single relation, called \verb|name|. The relation associates a \verb|NameEntry| with each name in the scope graph. The \verb|NameEntry| can be either a \verb|NType|, which stores the type of a name, or a \verb|NSubst|, which stores a name that has been substituted with a value. The Statix definition of these concepts is given below:
\begin{lstlisting}
constructors
    NType : Expr -> NameEntry
    NSubst : scope * Expr -> NameEntry
relations
    name : ID -> NameEntry
\end{lstlisting}

@@

We implemented beta-head reduction using a Krivine abstract machine~\cite{krivine}. The machine can head evaluate lambda expressions with a call-by-name semantics. This is a strategy under which the leftmost, outermost term is always reduced first~\cite{tapl}. It works by keeping a stack of all arguments that have not been applied yet. This turned out to be the more natural way of expressing this compared to substitution-based evaluation relation, which is an alternative we will discuss in section \ref{sec:coc-subst}.

In conventional dependently typed languages, evaluation is often done using De Bruijn indices, which use numbers to identify variables. However, we chose to use names rather than De Bruijn indices, because scope graphs work based on names. Using De Bruijn indices would also prevent us from using editor services that rely on \verb|.ref| annotations (which are Spoofax annotations that declare one name as being a use of another name that is a definition).

We need to define multiple predicates that will be used later for type checking. First, the primary predicate is \verb|betaReduceHead|, that takes a scoped expression and a stack of applications, and returns a head-normal expression. The scope acts as the environment from~\cite{krivine}, using \verb|NSubst| to store substitutions. All rules for \verb|betaReduceHead| are given in figure \ref{fig:beta-head-reduce-rules}. We use the syntax $\bhr{\scope{s_1}{e_1}}{\overline{p}}{\scope{s_2}{e_2}}$ to express \verb|betaReduceHead((s1, e1), ps) == (s2, e2)|. The $\overline{p}$ in this definition is the argument stack of the Krivine machine. The argument stack is a stack of scoped expressions, which are the arguments that are not yet paired with a matching function. Figure \ref{fig:beta-head-reduce-rules} contains the rules necessary for beta head reduction of the language. One predicate that is used for this is the \verb|rebuild| predicate, which takes a scoped expression and the stack of arguments (of the Krivine machine state) and converts it to an expression by adding \verb|FnDestruct|s. Its signature is:
\begin{lstlisting}
rebuild : (scope * Expr) * list((scope * Expr)) -> (scope * Expr)
\end{lstlisting}

@@

The inference rules in figures \ref{fig:beta-head-reduce-rules}, \ref{fig:beta-eq-rules}, and \ref{fig:type-check-rules} can be directly translated to Statix rules. For example, the rule for \verb|Let| bindings in figure \ref{fig:type-check-rules} is expressed like this in Statix:
\begin{lstlisting}
typeOfExpr(s, Let(n, v, b)) = typeOfExpr(s', b) :-
    typeOfExpr(s, v) == vt, sPutSubst(s, n, (s, v)) == s'.
\end{lstlisting}

@@

Although these rules look cleaner, they are more complicated to implement, requiring a separate relation to check if \verb|f| is a \verb|FnConstruct| or something else. For the pure calculus of constructions this still works quite well, but when adding inductive datatypes (chapter \ref{ch:datatypes}) the rules required become a lot more complex than those for a Krivine machine.
